### 2024-05-31 journal/analysis.tex
working on reviewer 2's comments

\begin{lemma}\label{dSearchTime}
The search that \op{FindResponse}$(b,i)$ does at line \ref{FRb} to find the index $b_e$ of the block in the root containing the $e$th enqueue takes $O(\log (\var{root}.\fld{blocks}[b_e].\size + \var{root}.\fld{blocks}[b-1].\size))$ steps.
\end{lemma}
\begin{proof}
Let the blocks in the root be $B_1, \ldots, B_\ell$.
The doubling search for $b_e$ takes $O(\log (b-b_e))$ steps,
so we prove $b - b_e \leq 2 \cdot B_{b_e}.\size + B_{b-1}.\size + 1$.
If $b \leq b_e+1$, then this is trivial, so assume  $b>b_e+1$.
%
As shown in Lemma \ref{linearCorrect}, the dequeue that calls \op{FindResponse} is in $B_b$ and is supposed to return an enqueue in $B_{b_e}$.
Thus, there can be at most $B_{b_e}.\size$ dequeues in 
$D(B_{b_e+1}) \cdots D(B_{b-1})$; otherwise in the sequential execution $L$,
all elements enqueued before the end of
$E(B_{b_e})$ would be dequeued before $D(B_b)$. 
Furthermore, by Lemma \ref{sizeCorrectness}, the size of the queue  after the prefix of $L$ corresponding to 
$B_1,\ldots,B_{b-1}$  is 
$B_{b-1}.\size \geq B_{b_e}.size + |E(B_{b_e+1})\cdots E(B_{b-1})| - |D(B_{b_e+1})\cdots D(B_{b-1})|$.
Thus, $|E(B_{b_e+1})\cdots E(B_{b-1})| \leq B_{b-1}.\size + |D(B_{b_e+1})\cdots D(B_{b-1})| \leq B_{b-1}.\size + B_{b_e}.\size$.
So, the total number of operations in $B_{b_e+1}, \ldots, B_{b-1}$ is at most
$B_{b-1}.\size + 2\cdot B_{b_e}.\size$.
Each of these $b-1-b_e$ blocks contains at least one operation, by Corollary \ref{blockNotEmpty}.
So, $b-1-b_e \leq B_{b-1}.\size + 2\cdot B_{b_e}.\size$.
\end{proof}

>>>>

\begin{lemma}\label{dSearchTime}
The search that \op{FindResponse}$(b,i)$ does at line \ref{FRb} to find the index $b_e$ of the block in the root containing the $e$th \enqueue\ takes $O(\log (\var{root}.\fld{blocks}[b_e].\size + \var{root}.\fld{blocks}[b-1].\size))$ steps.
\end{lemma}
\begin{proof}
Let the blocks in the root be $B_1, \ldots, B_\ell$.
The doubling search for $b_e$ takes $O(\log (b-b_e))$ steps,
so we prove $b - b_e \leq 2 \cdot B_{b_e}.\size + B_{b-1}.\size + 1$.
If $b \leq b_e+1$, then this is trivial, so assume  $b>b_e+1$.
%
As shown in Lemma \ref{linearCorrect}, the \dequeue\ that calls \op{FindResponse} is in $B_b$ and is supposed to return an \enqueue\ in $B_{b_e}$.
Thus, there can be at most $B_{b_e}.\size$ \dequeues\ in 
$D(B_{b_e+1}) \cdots D(B_{b-1})$; otherwise in the sequential execution $L$,
all elements enqueued before the end of
$E(B_{b_e})$ would be dequeued before $D(B_b)$. 
Furthermore, by Lemma \ref{sizeCorrectness}, the size of the queue  after the prefix of $L$ corresponding to 
$B_1,\ldots,B_{b-1}$  is 
$B_{b-1}.\size \geq B_{b_e}.size + |E(B_{b_e+1})\cdots E(B_{b-1})| - |D(B_{b_e+1})\cdots D(B_{b-1})|$.
Thus, $|E(B_{b_e+1})\cdots E(B_{b-1})| \leq B_{b-1}.\size + |D(B_{b_e+1})\cdots D(B_{b-1})| \leq B_{b-1}.\size + B_{b_e}.\size$.
So, the total number of operations in $B_{b_e+1}, \ldots, B_{b-1}$ is at most
$B_{b-1}.\size + 2\cdot B_{b_e}.\size$.
Each of these $b-1-b_e$ blocks contains at least one operation, by Corollary \ref{blockNotEmpty}.
So, $b-1-b_e \leq B_{b-1}.\size + 2\cdot B_{b_e}.\size$.
\end{proof}

### 2024-05-31 journal/boundedspace.tex
working on reviewer 2's comments

To maintain good amortized time, we periodically do a garbage collection (GC) phase.
If a \op{Refresh} on a node adds a block whose \var{index} is a multiple of $G=p^2\ceil{\log p}$, it does GC to remove obsolete blocks from the node's RBT.
To determine which blocks can be thrown away, we use a global array $\var{last}[1..p]$ where 
each process writes the index of the last block in the root
containing a null dequeue or an enqueue whose element it dequeued.
To perform GC, a process reads $\var{last}[1..p]$ and finds the maximum entry $m$.
Then, it helps complete every other process's pending dequeue 
by computing the dequeue's response and writing it in the block in the leaf that represents the dequeue.
Once this helping is complete, it follows from the FIFO property of the queue that elements enqueued 
in $\fld{root.blocks}[1..m-1]$ have all been dequeued, so GC can discard all subblocks of those.
Fortunately, there is an  RBT \op{Split} operation that can remove
these obsolete blocks from an RBT in logarithmic time \cite[Sec.~4.2]{Tar83}.

An operation $op$'s search of a RBT may fail to find the required block $B$ that has been removed 
by another process's GC phase.  If $op$
is a dequeue, $op$ must have been helped before $B$ was discarded, so $op$ can simply read its response from
its own leaf.  If $op$ is an enqueue, it can simply terminate.

>>>>

To maintain good amortized time, we periodically do a garbage collection (GC) phase.
If a \op{Refresh} on a node adds a block whose \var{index} is a multiple of $G=p^2\ceil{\log p}$, it does GC to remove obsolete blocks from the node's RBT.
To determine which blocks can be thrown away, we use a global array $\var{last}[1..p]$ where 
each process writes the index of the last block in the root
containing a null \dequeue\ or an \enqueue\ whose element it dequeued.
To perform GC, a process reads $\var{last}[1..p]$ and finds the maximum entry $m$.
Then, it helps complete every other process's pending \dequeue\ by 
computing the \dequeue's response and writing it in the block in the leaf that represents the \dequeue.
Once this helping is complete, it follows from the FIFO property of the queue that elements enqueued 
in $\fld{root.blocks}[1..m-1]$ have all been dequeued, so GC can discard all subblocks of those.
Fortunately, there is an  RBT \op{Split} operation that can remove
these obsolete blocks from an RBT in logarithmic time \cite[Sec.~4.2]{Tar83}.

An operation $op$'s search of a RBT may fail to find the required block $B$ that has been removed 
by another process's GC phase.  If $op$
is a \dequeue, $op$ must have been helped before $B$ was discarded, so $op$ can simply read its response from
its own leaf.  If $op$ is an \enqueue, it can simply terminate.

### 2024-05-31 journal/conclusion.tex
working on reviewer 2's comments

The approach used here to implement a lock-free queue 
could be applied to obtain other lock-free
data structures with a polylogarithmic step complexity.
For example, we can easily adapt the routines of the implementation in Section \ref{DescriptQ} 
to implement a restricted kind of vector data structure that stores a sequence and
provides three operations: \opa{Append}{e} to add an element \var{e} to the end of the sequence,
\opa{Get}{i} to read the \var{i}th element in the sequence, and
\opa{Index}{e} to compute the position of element \var{e} in the sequence.
Only the \op{Append} operations need to be propagated to the root of the ordering tree
since the other two operations do not affect the state of the object.
An \opa{Append}{e} is implemented like \opa{Enqueue}{e} in $O(\log p)$ steps.  
A \opa{Get}{i} is similar to \op{GetEnqueue}, taking $O(\log n + \log^2p)$ steps when the vector has $n$ elements.  
An \op{Index} is similar to \op{IndexDequeue} (except operating on enqueues instead of dequeues) and would take $O(\log p)$ steps if the argument is a pointer to the leaf block that contains the element~$e$.

>>>>

The approach used here to implement a lock-free queue 
could be applied to obtain other lock-free
data structures with a polylogarithmic step complexity.
For example, we can easily adapt the routines of the implementation in Section \ref{DescriptQ} 
to implement a restricted kind of vector data structure that stores a sequence and
provides three operations: \opa{Append}{e} to add an element \var{e} to the end of the sequence,
\opa{Get}{i} to read the \var{i}th element in the sequence, and
\opa{Index}{e} to compute the position of element \var{e} in the sequence.
Only the \op{Append} operations need to be propagated to the root of the ordering tree
since the other two operations do not affect the state of the object.
An \opa{Append}{e} is implemented like \opa{Enqueue}{e} in $O(\log p)$ steps.  
A \opa{Get}{i} is similar to \op{GetEnqueue}, taking $O(\log n + \log^2p)$ steps when the vector has $n$ elements.  
An \op{Index} is similar to \op{IndexDequeue} (except operating on \enqueues\ instead of \dequeues) and would take $O(\log p)$ steps if the argument is a pointer to the leaf block that contains the element~$e$.

### 2024-05-31 journal/correctness-GC.tex
working on reviewer 2's comments

\begin{customlemma}{\ref{successfulRefresh}$'$}\label{successfulRefreshGC}
Let $R$ be a call to \op{Refresh}(\var{v}) that performs a successful \op{CAS} on line \ref{casGC} (or terminates at line \ref{addOPGC}).
In the configuration after that CAS (or termination, respectively), \var{v} contains all operations that \var{v}'s children contained 
when $R$ executed line~\ref{refreshReadTGC}.
\end{customlemma}
\begin{proof}
Suppose \var{v}'s child (without loss of generality, $\var{v}.\fld{left}$) contained an operation $op$ 
when $R$ executed line \ref{refreshReadTGC}.
Let $i$ be the index of the block $B_\ell$ containing $op$ that was in \var{v}.\fld{left}'s RBT before 
$R$ executed line \ref{refreshReadTGC}.
We consider two cases.

@@

We say a block is \emph{finished} if 
\begin{itemize}
\item
it has been propagated to the root,
\item
the value of each enqueue contained in the block has either been returned by a dequeue
or written in the \fld{response} field of a dequeue, and
\item
each dequeue contained in the block has terminated or some process has written to the \fld{response} field in the leaf block that represents it.
\end{itemize}
Intuitively, once a block is finished, operations no longer need  the block
to compute responses to operations.
The following is an immediate consequence of the definition of finished and what it means for an operation to be contained in a block.

@@

%\here{Technically, following proof might have to be wrapped up in a big induction with the argument
%that outputs are linearized correctly, to argue that the matchup between enqueues and dequeues
%is the same as it would be in the linearization ordering.  (Check whether this is really needed.)
%Or alternatively, we could add an assumption that if every operation agrees with the linearization
%so far, then the invariant is preserved by each step.
%Then, when proving the linearization is correct, everything will work out.
%But just leave it as is for the submission, since this is a subtle point}
% NO, I DON'T THINK THIS IS NECESSARY

@@

For the base case, suppose $x$ is the root.
\op{SplitBlock} finds the maximum value $m$ in \var{last}, which is the index of some block
that contains an operation that is either a null dequeue or an enqueue whose value is the response for a dequeue that has been propagated to the root (since these are the only ways that an entry of \var{last} can be set to $m$).
By the FIFO property of queues, the values enqueued by enqueues in $\var{root}.\fld{blocks}[1..m-1]$
are all dequeued by operations that have already been installed in $\var{root}.\fld{blocks}$ before
the end of the \op{SplitBlock}.
Between the termination of $A$'s call to \op{SplitBlock} at line \ref{callSplitBlock}
and the \op{CAS} step $st$ after $A$ terminates,  $A$ helps all pending dequeues at line \ref{invokeHelp}.
Thus, after this helping (and before step $st$),
$\var{root}.\fld{blocks}[1..m-1]$ are all finished blocks.
Since \op{SplitBlock} returns \var{root}.\fld{blocks}[$m-1$], the claim is true.

@@

\er{edited this far}

\begin{lemma}\label{blocknotfound}
If a \op{Dequeue} operation fails to find a block in an RBT,
%during its calls to \op{IndexDequeue} or \op{GetEnqueue}, 
then it has been propagated to the root and its result has been written in the \fld{response} field of the leaf block it created.
If an \op{Enqueue} operation fails to find a block in an RBT, it has been propagated to the root.
\end{lemma}
\begin{proof}
Any block that GC removes from an RBT is finished, by Invariant \ref{inv::finished}, so if an 
\op{Enqueue} or \op{Dequeue} fails to find a block while it is propagating itself up to
the root (for example, during the \op{CreateBlock} routine), then a block containing
the operation itself has been removed from a RBT, so the operation has propagated to the root,
by the definition of finished.
Moreover, by Invariant \ref{inv::finished}, if the operation is a dequeue, then its result is in its \fld{response} field.

After propagation to the root, a \op{Dequeue} must access blocks that contain the dequeue,
as well as the enqueue whose value it will return (if it is not a null dequeue).
If any of those blocks have been removed, it follows from Invariant \ref{inv::finished} that the
\op{Dequeue}'s result is written in its \fld{response} field.
\end{proof}

@@

We first bound the size of RBTs.  Let $q_{max}$ be the maximum size of the queue at any time during the sequential execution given by the linearization $L$.
Recall that GC is done on a node every $G$ times its RBT is updated, and we chose $G$ to be $p^2\ceil{\log p}$.
Part of the proof of the following lemma is similar to the proof of Lemma \ref{dSearchTime}.
%We say the index $b$ \emph{is marked in node \var{v}} if some process has tested
%whether $b>\var{v}.\fld{last}[*]$ at line \ref{marktest1} or \ref{marktest2}.
%Intuitively, a block's index value is marked 
%when a dequeue passes through this block to find the enqueue
%whose argument it should return, or when a dequeue that returns null passes through the block.
%Thus, $b$ gets marked in \var{v} when $\var{v}.\fld{blocks}[b]$ 
%contains either an enqueue whose value is being dequeued or a \nl\ dequeue.
%If the index of block $B$ is marked in \var{v}, we also say $B$ is marked in \var{v}.

\begin{lemma}\label{boundAfterGC}
If the maximum \fld{index} in a node's RBT is a multiple of $G$, then
it contains at most $3q_{max}+5p+1$ blocks.
\end{lemma}
\begin{proof}
Consider the invocation $A$ of \op{AddBlock} that updates a node \var{v}'s RBT with the insertion of a block whose 
\fld{index} is a multiple of $G$.
Then, $A$ performs a GC phase.
Let $C$ be the configuration before $A$ invokes \op{SplitBlock} on line \ref{callSplitBlock}.
That call to \op{SplitBlock} recurses up to the root, where it computes $m$ by reading the \var{last}
array.
Let $L_1$ be the prefix of the linearization $L$ corresponding to blocks $1..m$ of the root.
Let $L_2$ be the next segment of the linearization corresponding to blocks $m+1..\ell$ of the root, where
$\ell$ is the last block added to the root's RBT before $C$.

@@

The number of enqueues in $L_2$ whose values are still in the queue at the end of $L_2$ is at most $q_{max}$.
If the value enqueued by any enqueue in $L_2$  is not still in the queue
at the end of $L_2$, then the dequeue in $L_2$ that dequeued that value must still be in progress
at $C$; otherwise the process that performed that dequeue would have set its \var{last} entry
to the index of the root block that contains the enqueue, which is greater than $m$, before $C$, contradicting
the fact that all values in the $last$ array at $C$ are less than or equal to $m$.
So, there are at most $p$ enqueues in $L_2$ whose values are still in the queue at the end of $L_2$.
Thus, there are at most $q_{max}+p$ enqueues in $L_2$.

If a dequeue in $L_2$ returns a non-null value in the sequential execution $L$,
then the value it returns was either in the queue at the end of $L_1$ or it was enqueued during $L_2$.
Thus, there are at most $q_{max}+(q_{max}+p)$ dequeues in $L_2$ that return non-null values.
Any dequeue in $L_2$ that returns a null value in the sequential execution $L$ must still be in progress
at $C$; otherwise the process that performed the dequeue would have set its \var{last} entry
to a value greater than the index of the root block that contains the dequeue prior to $C$, 
contradicting the definition of $m$.
So, there are at most $p$ null dequeues in $L_2$.
Thus, there are at most $2q_{max} + 2p$ dequeues in $L_2$.

@@

Next, we bound the number of operations in \var{v}'s blocks that are retained when $A$  sets
$T'\assign\op{Split}(T,s)$.
Since $T$ was read before $C$, any operation in $T$ is either in progress at $C$ or has been propagated to the root before $C$, by Lemma \ref{lem::appendExactlyOnce}.
Thus, there are at most $p$ operations in $T$ that do not appear in $L_1\cdot L_2$.
All the rest of the operations in blocks of $T'$ have been propagated to blocks $m..\ell$ of the root.
By Lemma \ref{blockSize}, There are at most $p$ operations in block $m$ of the root 
and we showed above that there are at most $3q_{max} + 3p$ in blocks 
$m+1..\ell$ of the root.
Thus, there are at most $3q_{max}+5p$ operations in blocks of $T'$.
Since each block is non-empty by Corollary \ref{blockNotEmpty}, $T'$ contains at most $3q_{max}+5p$ blocks, and one more 
block is inserted before $A$ sets \var{v}'s \fld{blocks} to the resulting RBT.
\end{proof}

>>>>

\begin{customlemma}{\ref{successfulRefresh}$'$}\label{successfulRefreshGC}
Let $R$ be a call to \op{Refresh}(\var{v}) that performs a successful \op{CAS} on line \ref{casGC} (or terminates at line \ref{addOPGC}).
Immediately after that CAS (or termination, respectively), \var{v} contains all operations that \var{v}'s children contained 
when $R$ executed line~\ref{refreshReadTGC}.
\end{customlemma}
\begin{proof}
Suppose \var{v}'s child (without loss of generality, $\var{v}.\fld{left}$) contained an operation $op$ 
when $R$ executed line \ref{refreshReadTGC}.
Let $i$ be the index of the block $B_\ell$ containing $op$ that was in \var{v}.\fld{left}'s RBT before 
$R$ executed line \ref{refreshReadTGC}.
We consider two cases.

@@

We say a block is \emph{finished} if 
\begin{itemize}
\item
it has been propagated to the root,
\item
the value of each \enqueue\ contained in the block has either been returned by a \dequeue\ or 
written in the \fld{response} field of a \dequeue, and
\item
each \dequeue\ contained in the block has terminated or some process has written to the \fld{response} field in the leaf block that represents it.
\end{itemize}
Intuitively, once a block is finished, operations no longer need  the block
to compute responses to operations.
The following is an immediate consequence of the definition of finished and what it means for an operation to be contained in a block.

@@

%\here{Technically, following proof might have to be wrapped up in a big induction with the argument
%that outputs are linearized correctly, to argue that the matchup between \enqueues\ and \dequeues\ 
%is the same as it would be in the linearization ordering.  (Check whether this is really needed.)
%Or alternatively, we could add an assumption that if every operation agrees with the linearization
%so far, then the invariant is preserved by each step.
%Then, when proving the linearization is correct, everything will work out.
%But just leave it as is for the submission, since this is a subtle point}
% NO, I DON'T THINK THIS IS NECESSARY

@@

For the base case, suppose $x$ is the root.
\op{SplitBlock} finds the maximum value $m$ in \var{last}, which is the index of some block
that contains an operation that is either a null \dequeue\ or an \enqueue\ whose value is the response for a \dequeue\ that has been propagated to the root (since these are the only ways that an entry of \var{last} can be set to $m$).
By the FIFO property of queues, the values enqueued by \enqueues\ in $\var{root}.\fld{blocks}[1..m-1]$
are all dequeued by operations that have already been installed in $\var{root}.\fld{blocks}$ before
the end of the \op{SplitBlock}.
Between the termination of $A$'s call to \op{SplitBlock} at line \ref{callSplitBlock}
and the \op{CAS} step $st$ after $A$ terminates,  $A$ helps all pending \dequeues\ at line \ref{invokeHelp}.
Thus, after this helping (and before step $st$),
$\var{root}.\fld{blocks}[1..m-1]$ are all finished blocks.
Since \op{SplitBlock} returns \var{root}.\fld{blocks}[$m-1$], the claim is true.

@@

\begin{lemma}\label{blocknotfound}
If a \op{Dequeue} operation fails to find a block in an RBT,
%during its calls to \op{IndexDequeue} or \op{GetEnqueue}, 
then it has been propagated to the root and its result has been written in the \fld{response} field of the leaf block it created.
If an \op{Enqueue} operation fails to find a block in an RBT, it has been propagated to the root.
\end{lemma}
\begin{proof}
Any block that GC removes from an RBT is finished, by Invariant \ref{inv::finished}, so if an 
\op{Enqueue} or \op{Dequeue} fails to find a block while it is propagating itself up to
the root (for example, during the \op{CreateBlock} routine), then a block containing
the operation itself has been removed from a RBT, so the operation has propagated to the root,
by the definition of finished.
Moreover, by Invariant \ref{inv::finished}, if the operation is a \dequeue, then its result is in its \fld{response} field.

After propagation to the root, a \op{Dequeue} must access blocks that contain the \dequeue,
as well as the \enqueue\ whose value it will return (if it is not a null \dequeue).
If any of those blocks have been removed, it follows from Invariant \ref{inv::finished} that the
\op{Dequeue}'s result is written in its \fld{response} field.
\end{proof}

@@

We first bound the size of RBTs.  Let $q_{max}$ be the maximum size of the queue at any time during the sequential execution given by the linearization $L$.
Recall that GC is done on a node every $G$ times its RBT is updated, and we chose $G$ to be $p^2\ceil{\log p}$.
Part of the proof of the following lemma is similar to the proof of Lemma \ref{dSearchTime}.
%We say the index $b$ \emph{is marked in node \var{v}} if some process has tested
%whether $b>\var{v}.\fld{last}[*]$ at line \ref{marktest1} or \ref{marktest2}.
%Intuitively, a block's index value is marked 
%when a \dequeue\ passes through this block to find the \enqueue
%whose argument it should return, or when a \dequeue\ that returns null passes through the block.
%Thus, $b$ gets marked in \var{v} when $\var{v}.\fld{blocks}[b]$ 
%contains either an \enqueue whose value is being dequeued or a \nl\ \dequeue.
%If the index of block $B$ is marked in \var{v}, we also say $B$ is marked in \var{v}.

\begin{lemma}\label{boundAfterGC}
If the maximum \fld{index} in a node's RBT is a multiple of $G$, then
it contains at most $3q_{max}+5p+1$ blocks.
\end{lemma}
\begin{proof}
Consider the invocation $A$ of \op{AddBlock} that updates a node \var{v}'s RBT with the insertion of a block whose 
\fld{index} is a multiple of $G$.
Then, $A$ performs a GC phase.
Let \rev{$t$ be the time} when $A$ invokes \op{SplitBlock} on line \ref{callSplitBlock}.
That call to \op{SplitBlock} recurses up to the root, where it computes $m$ by reading the \var{last}
array.
Let $L_1$ be the prefix of the linearization $L$ corresponding to blocks $1..m$ of the root.
Let $L_2$ be the next segment of the linearization corresponding to blocks $m+1..\ell$ of the root, where
$\ell$ is the last block added to the root's RBT before $t$.

@@

The number of \enqueues\ in $L_2$ whose values are still in the queue at the end of $L_2$ is at most $q_{max}$.
If the value enqueued by any \enqueue\ in $L_2$  is not still in the queue
at the end of $L_2$, then the \dequeue\ in $L_2$ that dequeued that value must still be in progress
at time $t$; otherwise the process that performed that \dequeue\ would have set its \var{last} entry
to the index of the root block that contains the \enqueue, which is greater than $m$, before $t$, contradicting
the fact that all values in the $last$ array at $t$ are less than or equal to $m$.
So, there are at most $p$ \enqueues\ in $L_2$ whose values are still in the queue at the end of $L_2$.
Thus, there are at most $q_{max}+p$ \enqueues\ in $L_2$.

If a \dequeue\ in $L_2$ returns a non-null value in the sequential execution $L$,
then the value it returns was either in the queue at the end of $L_1$ or it was enqueued during $L_2$.
Thus, there are at most $q_{max}+(q_{max}+p)$ \dequeues\ in $L_2$ that return non-null values.
Any \dequeue\ in $L_2$ that returns a null value in the sequential execution $L$ must still be in progress
at $t$; otherwise the process that performed the \dequeue\ would have set its \var{last} entry
to a value greater than the index of the root block that contains the \dequeue\ prior to $t$, 
contradicting the definition of $m$.
So, there are at most $p$ null \dequeues\ in $L_2$.
Thus, there are at most $2q_{max} + 2p$ \dequeues\ in $L_2$.

@@

Next, we bound the number of operations in \var{v}'s blocks that are retained when $A$  sets
$T'\assign\op{Split}(T,s)$.
Since $T$ was read before $t$, any operation in $T$ is either in progress at $t$ or has been propagated to the root before $t$, by Lemma \ref{lem::appendExactlyOnce}.
Thus, there are at most $p$ operations in $T$ that do not appear in $L_1\cdot L_2$.
All the rest of the operations in blocks of $T'$ have been propagated to blocks $m..\ell$ of the root.
By Lemma \ref{blockSize}, There are at most $p$ operations in block $m$ of the root 
and we showed above that there are at most $3q_{max} + 3p$ in blocks 
$m+1..\ell$ of the root.
Thus, there are at most $3q_{max}+5p$ operations in blocks of $T'$.
Since each block is non-empty by Corollary \ref{blockNotEmpty}, $T'$ contains at most $3q_{max}+5p$ blocks, and one more 
block is inserted before $A$ sets \var{v}'s \fld{blocks} to the resulting RBT.
\end{proof}
